---
title: "Reproducible Course Project 1"
output: html_document
---

## Loading the needed libraries    
Below are the libraries needed:  

```{r, echo=TRUE}
library(dplyr)
```

## Loading and preprocessing the data  
Below is code that is needed to:  

1. Load the data  

```{r, echo=TRUE}
activity <- read.csv("activity.csv")
```

2. Process/transform the data  

```{r, echo=TRUE}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?  
For this part of the assignment, I am ignoring the missing values in the dataset.  

1. Calculate the total number of steps taken per day  

```{r, echo=TRUE}
dailyTotalSteps <- activity %>% group_by(date) %>% summarise(total = sum(steps))
```

2. Make a histogram of the total number of steps taken each day  

```{r daily-total, echo=TRUE, fig.path="figure/"}
with(dailyTotalSteps, hist(total, main = "Daily Total"))
```

3. Calculate the mean and median of the total number of steps taken per day  

```{r, echo=TRUE}
mean(dailyTotalSteps$total, na.rm = TRUE)
median(dailyTotalSteps$total, na.rm = TRUE)
```

## What is the average daily activity pattern?  

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r daily-average, fig.path="figure/", echo=TRUE}
intervalAveSteps <- activity %>% group_by(interval) %>% summarise(average = mean(steps, na.rm = TRUE))
with(intervalAveSteps, plot(interval, average, type = "l", main = "Daily Average"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  

```{r, echo=TRUE}
intervalAveSteps$interval[intervalAveSteps$average == max(intervalAveSteps$average)]
```

## Imputing missing values  

1. Calculate and report the total number of missing values in the dataset  

```{r, echo=TRUE}
sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset.  

```{r, echo=TRUE}
imputeStep <- function(step, interval){
  if(is.na(step)) {
    round(intervalAveSteps$average[intervalAveSteps$interval == interval])
  } else {
    step
  }
}
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.  

```{r, echo=TRUE}
newActivity <- mutate(activity, steps = mapply(imputeStep, steps, interval))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.  


```{r daily-imputed-total, fig.path="figure/", echo=TRUE}
newDailyTotalSteps <- newActivity %>% group_by(date) %>% summarise(total = sum(steps))
with(newDailyTotalSteps, hist(total, main = "Daily Imputed Total"))
mean(newDailyTotalSteps$total, na.rm = TRUE)
median(newDailyTotalSteps$total, na.rm = TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?  

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.  

```{r, echo=TRUE}
makeFactor <- function(date){
  factor(weekdays(date) %in% c("Saturday","Sunday"), levels = c(TRUE,FALSE), labels = c("weekend","weekday"))
}
newActivity <- mutate(newActivity, dayOfWeek = makeFactor(date))
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).  

```{r weekday-average, fig.path="figure/", echo=TRUE}
weekdayActivity <- newActivity %>% filter(dayOfWeek == "weekday") %>% group_by(interval) %>% summarise(average = mean(steps))

weekendActivity <- newActivity %>% filter(dayOfWeek == "weekend") %>% group_by(interval) %>% summarise(average = mean(steps))

par(mfrow = c(1, 2))
with(weekdayActivity, plot(interval, average, type = "l", main = "Weekday Average"))
with(weekendActivity, plot(interval, average, type = "l", main = "Weekend Average"))
```




